### 2025-02-27 paper/conclusions.tex
Pass throught the paper.

The paper demonstrates a technique of developing high performance
applications with strong correctness guarantees, including the absence
of out-of-bound indexing, certain functions being inverses, as well as
well-scopedness and well-typedness of our DSL.

@@

A lot of pieces that we have developed in this paper can be reused in
other numerical applications. However, there are many more
opportunities that we did not explore. For example, one can prove the
correctness of optimisations, relating evaluation of optimised an
non-optimised expressions. We can provide more guarantees when we run
extraction, \eg{} we can formalise some aspects of the backend
language and relate them to our DSL. As for the DSL itself, the exact
choice of supported primitives, and its implications for extracting
high performance code or performing optimisation or transformation,
remains an interesting question.

>>>>

The paper demonstrates a technique of developing high performance
applications with strong correctness guarantees, including the absence
of out-of-bound indexing, certain functions being inverses,
well-scopedness and well-typedness of the embedded DSL and
semantics-preserving optimisations.

@@

A lot of pieces that we have developed in this paper can be reused in
other numerical applications. However, there are many more
opportunities that we did not yet explore. We can formalise other
correctness criteria for our specification spanning from verified
extraction to functional correctness of the actual application such
as CNN robustness.
Choosing DSL primitives so that specificational clarity and
performance requirements are at balance is an interesting question.
Guaranteeing asymptotic efficiency of the AD implementation is another
interesting research direction.  Currently we rely on
optimisations which is known to be a fragile technique, despite its efficiency 
in many cases.  Identifying a normal form for the optimisations would
be a more fundamental approach.  Richer languages with more
control flow may require us choosing more powerful
encoding for the adjoint structures.

### 2025-02-27 paper/relatedwork.tex
Pass throught the paper.

\subsection{Automatic Differentiation}

@@

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
\todo[inline]{reviewer 3 (2024) notices: This paper does not really answer this question besides enforcing consistency of shapes and accesses? Besides, final extraction could break these properties?  this is correct, we have to clarify this.}
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.

@@

\subsection{Verified high performance computing}

@@

\subsection{Type systems}

>>>>

\paragraph{Automatic Differentiation}

@@

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Note that while our work
does not give a formal proof that AD implementation is correct (it is
not clear how to formulate correcntess criteria for it), the entire
implementatuion of AD is about 20 lines of code, so it is easy to
understand and modify.

@@

\paragraph{Verified high performance computing}

@@

\paragraph{Type systems}
